Progression Features & OAC

### 2018-05-02 features-OAC.Rmd
Moved shiny app to own repo

library(pander)
library(gridExtra)
library(plyr)
library(ggplot2)
library(reshape2)
library(GenomicRanges)
library(ggrepel)
library(ggfortify)
library(readxl)

@@

## Lists the sequenced patients
sequenced = read_excel('~/Documents/OneDrive/OneDrive - MRC Cancer Unit at University of Cambridge/skillcoyne/OC Data/WGS_Phase 1 complete - cleaned.xlsx', trim_ws=T)
sequenced = subset(sequenced, grepl('^OC', OCCAMS_ID)  )[,c(1:8)]
sequenced$Sample_type = as.factor(tolower(sequenced$Sample_type))
sequenced = subset(sequenced, grepl('barret|normal|tumour', Sample_type))

@@

# OCCAMS ID to Illumina ID
paired = read.table('~/Documents/OneDrive/MRC Cancer Unit at University of Cambridge/icgc group - icgc_shared/icgc-group/batch_progress/reports/g37/PairedOverviewSummary_g37_bwa.txt', header=T, sep='\t', stringsAsFactors = F)
paired = subset(paired, aligned == '*')
pairedg37 = subset(paired, OCCAMS %in% unique(sequenced$OCCAMS_ID), select=c('OCCAMS','SampleId'))
length(unique(pairedg37$OCCAMS))

paired = read.table('~/Documents/OneDrive/MRC Cancer Unit at University of Cambridge/icgc group - icgc_shared/icgc-group/batch_progress/reports/g1k//PairedOverviewSummary_g1k_bwa.txt', header=T, sep='\t', stringsAsFactors = F)
paired = subset(paired, aligned == '*')
pairedg1k = subset(paired, OCCAMS %in% unique(sequenced$OCCAMS_ID), select=c('OCCAMS','SampleId'))

paired = rbind(pairedg37, pairedg1k)
paired = unique(paired)

length(unique(paired$OCCAMS))

paired[] = lapply(paired[], strip.whitespace)
length(unique(paired$OCCAMS))

sp = do.call(rbind, lapply(paired$SampleId, function(x) unlist(strsplit(x, '_vs_'))))
colnames(sp) = c('tb.sample', 'normal')

paired = cbind(paired, sp)

barretts.pairs = subset(paired, tb.sample %in% be.seq$Illumina_Barcode)

paired = paired[-which(paired$tb.sample %in% be.seq$Illumina_Barcode),]

# ------------------------------ #
# Clinical data
load('~/Data/ICGC/OAC/clinical/labkey_20172706.Rdata', verbose=T)
demo = occams$patients
cols = c(grep('Reason',colnames(demo), value=T),
         grep('Planned',colnames(demo),value=T),
         'FE.EndPoint','RP.Tumor.Size','ST.MainSurgery','RP.HistoryOfNeoAdjuvantTherapy',
         'TR.PalliativeTreatmentModality','TR.CurativeTreatmentModality')
demo[cols] = lapply(demo[cols], as.null)
#demo = oac.demo[unique(sequenced$OCCAMS_ID),]
# ------------------------------ #

@@

demo[rownames(ba[changed,]), 'RP.BarrettsAdjacent'] = ba[changed,'caitrona']
# ------------------------------ #

# Copy number data - tumour
read.caveman<-function(s) {
  files = sample.list[[s]]
  if (length(files) < 2) {
    warning(s)
    stop()
  }
  stats = read.table(grep('samplestatistics.csv', files, value=T), row.names=1)

  caveman = read.csv(grep('copynumber.caveman.csv$', files, value=T))
  caveman$ploidy = stats['Ploidy',]
  caveman$rho = stats['rho',]
  caveman$contamination = stats['NormalContamination',]

  caveman$Illumina.Barcode.full = s
  caveman
}

# Tumours
dir = '~/Data/ICGC/BE_OAC_Cohort/tumour/data/ascat'
files = list.files(dir, full.names=T, recursive=T)
samples = list.files(dir, full.names=F)

sample.list = lapply(samples, function(x) {
  grep(x, files, value=T)
} )
names(sample.list) = samples

cnv = do.call(rbind, lapply(samples, read.caveman))
totalCNV = merge(cnv, paired, by.x='Illumina.Barcode.full', by.y='SampleId')

demo.cols = c('StudySubjectID','RP.TStage.PrimaryTumour','RP.BarrettsAdjacent','Weeks.Survival','RP.TumourDifferentiation','DI.ageAtDiagnosis','DI.PatientGender','FE.Patient.Died')
totalCNV = merge(totalCNV, demo[,demo.cols], by.y='StudySubjectID', by.x='OCCAMS')

oacCNV = makeGRangesFromDataFrame(totalCNV, keep.extra.columns = T)

# Barretts ADJACENT SAMPLES
dir = '~/Data/ICGC/BE_OAC_Cohort/BA/data/ascat'
files = list.files(dir, full.names=T, recursive=T)
samples = list.files(dir, full.names=F)

sample.list = lapply(samples, function(x) {
  grep(x, files, value=T)
} )
names(sample.list) = samples

cnv = do.call(rbind, lapply(samples, read.caveman))
beCNV = merge(cnv, barretts.pairs, by.x='Illumina.Barcode.full', by.y='SampleId')
beCNV = merge(beCNV, demo[,demo.cols], by.y='StudySubjectID', by.x='OCCAMS')
beCNV = makeGRangesFromDataFrame(beCNV, keep.extra.columns = T)
# ------------------------------ #

oac.demo = subset(demo, StudySubjectID %in% unique(oacCNV$OCCAMS))
be.demo = subset(demo, StudySubjectID %in% unique(beCNV$OCCAMS))

## Shallow WGS data 
load('~/Data/Ellie/Analysis/5e06/loo.Rdata', verbose=T)
swgsPredictions = pg.samp
rm(pg.samp)

load('~/Data/Ellie/Analysis/5e06/all.pt.alpha.Rdata', verbose=T)
rawSWGSdf = dysplasia.df
a = '0.9'
coefs = coefs[[a]]; model = models[[a]]

hazards = apply( exp(t(rawSWGSdf[, rownames(coefs)])*coefs[,1]), 1, function(x) { sd(x)/mean(x) })
rm(dysplasia.df,fits,plots,performance.at.1se)
# ------------------------------ #

@@

analysis.files = list.files('~/Data/Ellie/Analysis', full.names=T)
load(grep('Training_patients.Rdata', analysis.files, value=T), verbose=T)

labels = unlist(sapply(patient.data, function(df) {
  df$info = arrange(df$info, Endoscopy.Year, Pathology)
  label = as.integer(df$info$Status == 'P') #as.integer(df$info$Pathology %in% c('HGD', 'IMC'))
  names(label) = df$info$Samplename
  return(label)
}))
names(labels) = sub('.*\\.', '',  names(labels))

labels = labels[which(names(labels) %in% rownames(rawSWGSdf))]

rm(patient.data)
```

@@

Load the coefficients from the selected model and merge with the copynumber regions identified in (some) of the OAC patients
```{r, echo=F}
regions = do.call(rbind.data.frame, c(strsplit(grep('\\d+', ch$label, value=T), ':|-'), stringsAsFactors=F))
colnames(regions) = c('chr','start','stop')
regions[colnames(regions)] = lapply(regions[colnames(regions)], function(x) as.numeric(x))

@@

```{r}
# positive coefficients were related to gains in progressors
gains = subset(ch, coef > 0)

gain = do.call(rbind, lapply(gains$label, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values > 1)])
}))

t.test(gain[,1], gain[,2])

@@

loss = do.call(rbind, lapply(losses$label, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values < -1)])
}))

t.test(loss[,1], loss[,2])

@@

randomGains = do.call(rbind, lapply(sample.reg, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values > 1)])
}))

t.test(randomGains[,1], randomGains[,2])

randomLosses = do.call(rbind, lapply(sample.reg, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values < -1)])
}))

t.test(randomLosses[,1], randomLosses[,2])

However, while there appears to be no difference in the gains found in NP for randomly selected regions vs the coefficients:

pander(wilcox.test(gain[,'0'], randomGains[,'0']), caption="Non progressors, coeffiecents vs randomly selected")

pander(wilcox.test(loss[,'0'], randomLosses[,'0']), caption="Non progressors, coeffiecents vs randomly selected")

There is a significant difference in the selected regions for progressors. Again this is mostly a sanity check:

pander(wilcox.test(gain[,'1'], randomGains[,'1']), caption="Progressors, coeffiecents vs randomly selected gains")

pander(wilcox.test(loss[,'1'], randomLosses[,'1']), caption="Progressors, coeffiecents vs randomly selected losses")

@@

```{r, fig.height=16, fig.width=16}
plot.gains<-function(df) {
  df = lapply(df, function(x) {
    x = arrange(x, -width)
    x$order = factor(1:nrow(x))
    return(x)
  })

  m = do.call(rbind.data.frame, df)
  m = m %>% group_by(seqnames) %>% ungroup() %>% arrange(segment, seqnames, width, start) %>% mutate(segorder = row_number())
m$isWGD = as.factor(m$isWGD)

@@

# WGD
oacCNV$isWGD = oacCNV$ploidy > 3.2 
beCNV$isWGD = beCNV$ploidy > 3.2

@@

## gains in patients -- issues with WGD samples?
gains = lapply(swgRegions[gr], gain.for, cnv=oacCNV)
names(gains) = as.character(swgRegions[gr])
plot.gains(gains) + labs(title='Gains in OAC samples')

be.gains = lapply(swgRegions[gr], gain.for, cnv=oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'yes')])
names(be.gains) = as.character(swgRegions[gr])
plot.gains(be.gains[sapply(be.gains, nrow) > 0]) + labs(title='Gains in BE Adjacent samples')

non.be.gains = lapply(swgRegions[gr], gain.for, cnv=oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'no')])
names(non.be.gains) = as.character(swgRegions[gr])
plot.gains(non.be.gains[sapply(non.be.gains, nrow) > 0]) + labs(title='Gains in NO BE Adjacent samples')

@@

# Per patient
gfreq = sort(sapply(gains, function(x) length(unique(x$OCCAMS)))/length(unique(sequenced$OCCAMS_ID)))
ggplot(melt(as.matrix(gfreq)), aes(Var1, value)) + geom_col(fill='darkblue') + ylim(0,0.5) + coord_flip() + 
  labs(x='', y='Frequency', title='Gain frequency in OAC') + theme_bw() + theme(text=element_text(size=14))

gfreq1 = sort(sapply(be.gains, function(x) length(unique(x$OCCAMS)))/length(unique(oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'yes')]$OCCAMS)))
gfreq1 = gfreq1[names(gfreq)]

gfreq2 = sort(sapply(non.be.gains, function(x) length(unique(x$OCCAMS)))/length(unique(oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'no')]$OCCAMS)))
gfreq2 = gfreq2[names(gfreq)]

@@

```{r, fig.height=12, fig.width=12}
plot.losses<-function(df) {
  if (!is.data.frame(df)) {
    df = lapply(df, function(x) {
      if(nrow(x) <= 0) return(NULL)
      x = arrange(x, -width)
      x$order = factor(1:nrow(x))
      return(x)
    })
    m = do.call(rbind.data.frame, df)
  } else {
    m = df
    m = arrange(m, -width)
    m$order = factor(1:nrow(m))
  }
  #m$segnames = ordered(m$seqnames, levels=c(1:22))

  m = m %>% group_by(seqnames) %>% ungroup() %>% arrange(segment, seqnames, width, start) %>% mutate(segorder = row_number())
  m$isWGD = as.factor(m$isWGD)
  m$CN = as.factor(m$Total_CN)
  #m = transform(m, segment=reorder(segment, segnames))
  p = ggplot(data=m, aes(x=order, y=start))

@@

losses = lapply(swgRegions[lr], loss.for, oacCNV)
names(losses) = as.character(swgRegions[lr])
plot.losses(losses)

be.losses = lapply(swgRegions[lr], loss.for, oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'yes')])
names(be.losses) = as.character(swgRegions[lr])
plot.losses(be.losses) + labs(title="Losses in (YES) BE Adjacent samples")

non.be.losses = lapply(swgRegions[lr], loss.for, cnv=oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'no')])
names(non.be.losses) = as.character(swgRegions[lr])
plot.losses(non.be.losses) + labs(title='Losses in (NO) BE Adjacent samples')

@@

lfreq = sort(sapply(losses, function(x) length(unique(x$OCCAMS)))/length(unique(sequenced$OCCAMS_ID)))
ggplot(melt(as.matrix(lfreq)), aes(Var1, value)) + geom_col(fill='darkgreen') + ylim(0,1) + coord_flip() + 
  labs(x='', y='Frequency', title='Loss frequency in OAC') + theme_bw() + theme(text=element_text(size=14))

lfreq1 = sapply(be.losses, function(x) length(unique(x$OCCAMS)))/length(unique(oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'yes')]$OCCAMS))
lfreq1 = lfreq1[names(lfreq)]

lfreq2 = sort(sapply(non.be.losses, function(x) length(unique(x$OCCAMS)))/length(unique(oacCNV[which(oacCNV$RP.BarrettsAdjacent == 'no')]$OCCAMS)))
lfreq2 = lfreq2[names(lfreq)]

@@

```{r fig.height=6,fig.width=8}

## Note it's a subset of the 321 patients (66) that have this
chr17 = get.chr.lengths()[17,]

#chr17losses = lapply(makeGRangesFromDataFrame( cbind.data.frame('start'=0, 'end'=chr17$chr.length, 'chr'='17') ), loss.for)[[1]]
#m = arrange(chr17losses, width, start)

plot.losses(losses$`17:1-4776188`) + geom_hline(data=chr17, aes(yintercept=(chr.cent-cent.gap)), color='red', linetype='dashed') +
  labs(y='chr17 position', x='Patient', title='Chr17 Losses in OAC Patients', subtitle=paste(round(length(unique(losses$`17:1-4776188`$OCCAMS))/length(unique(oacCNV$OCCAMS)), 2)*100, '% of OAC patients', sep=''))

@@

So is there a difference between the gains/losses in the selected regions?  
```{r}
wilcox.test(sapply(gains, nrow), sapply(losses, nrow))
one.way.aov(sapply(gains, nrow), sapply(losses, nrow))
```

@@

```{r}
randomGains = lapply(sample.regions, gain.for, cnv=oacCNV)
names(randomGains) = as.character(sample.regions)

@@

# Maybe a difference in # gains in the selected features vs random
wilcox.test(sapply(randomGains, nrow), sapply(gains, nrow))

wilcox.test(sapply(randomGains, nrow), sapply(be.gains,nrow))
wilcox.test(sapply(randomGains, nrow), sapply(non.be.gains,nrow))

# Not a difference here though
wilcox.test(sapply(gains, function(df) length(unique(df$OCCAMS))), 
            sapply(randomGains, function(df) length(unique(df$OCCAMS))))

df = rbind(cbind.data.frame('counts'=sapply(gains, function(df) length(unique(df$OCCAMS))), 'type'='selected') , 
      cbind.data.frame('counts'=sapply(randomGains, function(df) length(unique(df$OCCAMS))), 'type'='random'))
ggplot(df, aes(type, counts, group=type, fill=type)) + geom_boxplot() + geom_jitter(width=0.3) + labs(title='Patient counts by region') + plot.theme

@@

```{r}
randomLosses = lapply(sample.regions, loss.for, cnv=oacCNV)
names(randomLosses) = as.character(sample.regions)

@@

# No difference in # losses in the selected features
wilcox.test(sapply(randomLosses, nrow), sapply(losses, nrow))

wilcox.test(sapply(randomLosses, nrow), sapply(be.losses,nrow))
wilcox.test(sapply(randomLosses, nrow), sapply(non.be.losses,nrow))

# No difference between the two
df = rbind(cbind.data.frame('counts'=sapply(losses, function(df) length(unique(df$OCCAMS))), 'type'='selected') , 
      cbind.data.frame('counts'=sapply(randomLosses, function(df) length(unique(df$OCCAMS))), 'type'='random'))
ggplot(df, aes(type, counts, group=type, fill=type)) + geom_boxplot() + geom_jitter(width=0.3) + labs(title='Loss counts by patient') + plot.theme

@@

## Cox

>>>>

library(pander)
library(gridExtra)
library(ggplot2)
library(reshape2)
library(GenomicRanges)
library(ggrepel)
library(ggfortify)
library(readxl)
library(plyr)
library(dplyr)

@@

## Lists the sequenced patients
sequenced = read_xlsx('~/Documents/OneDrive/OneDrive - MRC Cancer Unit at University of Cambridge/skillcoyne/Documents/Sequencing Batch Info/occams_sequenced_20180426.xlsx')
sequenced = subset(sequenced, grepl('^OC', OCCAMS_ID) )
sequenced$Sample_type = as.factor(tolower(sequenced$Sample_type))
sequenced = subset(sequenced, grepl('barret|normal|tumour', Sample_type))

@@

# # OCCAMS ID to Illumina ID
# paired = read.table('~/Documents/OneDrive/MRC Cancer Unit at University of Cambridge/icgc group - icgc_shared/icgc-group/batch_progress/reports/g37/PairedOverviewSummary_g37_bwa.txt', header=T, sep='\t', stringsAsFactors = F)
# paired = subset(paired, aligned == '*')
# pairedg37 = subset(paired, OCCAMS %in% unique(sequenced$OCCAMS_ID), select=c('OCCAMS','SampleId'))
# length(unique(pairedg37$OCCAMS))
# 
# paired = read.table('~/Documents/OneDrive/MRC Cancer Unit at University of Cambridge/icgc group - icgc_shared/icgc-group/batch_progress/reports/g1k//PairedOverviewSummary_g1k_bwa.txt', header=T, sep='\t', stringsAsFactors = F)
# paired = subset(paired, aligned == '*')
# pairedg1k = subset(paired, OCCAMS %in% unique(sequenced$OCCAMS_ID), select=c('OCCAMS','SampleId'))
# 
# paired = rbind(pairedg37, pairedg1k)
# paired = unique(paired)
# 
# length(unique(paired$OCCAMS))
# 
# paired[] = lapply(paired[], strip.whitespace)
# length(unique(paired$OCCAMS))
# 
# sp = do.call(rbind, lapply(paired$SampleId, function(x) unlist(strsplit(x, '_vs_'))))
# colnames(sp) = c('tb.sample', 'normal')
# 
# paired = cbind(paired, sp)
# 
# barretts.pairs = subset(paired, tb.sample %in% be.seq$Illumina_Barcode)
# 
# paired = paired[-which(paired$tb.sample %in% be.seq$Illumina_Barcode),]

# ------------------------------ #
# Clinical data
load('~/Data/ICGC/OAC/clinical/labkey_20180111.Rdata', verbose=T)
demo = occams$patients
cols = c(grep('Reason',colnames(demo), value=T),
         grep('Planned',colnames(demo),value=T),
         'FE.EndPoint','RP.Tumor.Size','ST.MainSurgery','RP.HistoryOfNeoAdjuvantTherapy',
         'TR.PalliativeTreatmentModality','TR.CurativeTreatmentModality')
demo[cols] = lapply(demo[cols], as.null)
#demo = oac.demo[unique(sequenced$OCCAMS_ID),]
# ------------------------------ #

@@

demo[rownames(ba[changed,]), 'RP.BarrettsAdjacent'] = ba[changed,'caitrona']

# ------------------------------ #

# Tumours
# there are some duplicates in the patient samples so I've selected one tumor for each of those. That means that there are a few less samples to be analysed than the files initially might suggest
tumors = subset(sequenced, Sample_type == 'tumour')

cnv = read.table('~/Data/ICGC/BE_OAC_Cohort/tumour/data/ascat/totalCNV.txt', sep='\t', header=T, stringsAsFactors=F)

barcodes = do.call(rbind, strsplit(cnv$Illumina.Barcode, '_vs_'))
colnames(barcodes) = c('Tumor.Barcode','Normal.Barcode')

cnv = cbind(cnv, barcodes)
cnv[colnames(barcodes)] = lapply(cnv[colnames(barcodes)], as.character)
length(unique(cnv$Tumor.Barcode))

#subset(tumors, Illumina_Barcode %in% setdiff(cnv$Tumor.Barcode, tumors$Illumina_Barcode))

cnv = base::merge(cnv, tumors, by.x='Tumor.Barcode', by.y='Illumina_Barcode')

tumorcnv = makeGRangesFromDataFrame(cnv, keep.extra.columns = T)

# Not a problem at the moment
#which(!cnv$Tumor.Barcode %in% tumors$Illumina_Barcode)

# Barretts ADJACENT SAMPLES
becnv = read.table('~/Data/ICGC/BE_OAC_Cohort/BA/data/ascat/totalCNV.txt', sep='\t', header=T, stringsAsFactors=F)
length(unique(becnv$Illumina.Barcode))

barcodes = do.call(rbind, strsplit(becnv$Illumina.Barcode, '_vs_'))
colnames(barcodes) = c('Tumor.Barcode','Normal.Barcode')

becnv = cbind(becnv, barcodes)

becnv[colnames(barcodes)] = lapply(becnv[colnames(barcodes)], as.character)

becnv = base::merge(becnv, subset(sequenced, grepl('barr', Sample_type)) , by.x='Tumor.Barcode', by.y='Illumina_Barcode') 
#becnv = merge(becnv, demo[,cols], by.y='StudySubjectID', by.x='OCCAMS')
becnv = makeGRangesFromDataFrame(becnv, keep.extra.columns = T)
# ------------------------------ #

#oac.demo = subset(demo, StudySubjectID %in% unique(tumorcnv$OCCAMS))
#be.demo = subset(demo, StudySubjectID %in% unique(becnv$OCCAMS))

## Shallow WGS data 
load('~/Data/Ellie/Analysis/5e6_arms_all_exAHM0320/loo.Rdata', verbose=T)
swgsPredictions = pg.samp
rm(pg.samp, plots, coefs, fits)

load('~/Data/Ellie/Analysis/5e6_arms_all_exAHM0320/all.pt.alpha.Rdata', verbose=T)
rawSWGSdf = dysplasia.df
a = '0.9'
coefs = coefs[[a]]; model = models[[a]]

# CV(RR)
hazards = apply( exp(t(rawSWGSdf[, rownames(coefs)])*coefs[,1]), 1, function(x) { sd(x)/mean(x) })
rm(dysplasia.df,models,plots,performance.at.1se)
# ------------------------------ #

@@

@@

Load the coefficients from the selected model and merge with the copynumber regions identified in (some) of the OAC patients
```{r, echo=F, warning=F, message=F}
regions = do.call(rbind.data.frame, c(strsplit(grep('\\d+', ch$label, value=T), ':|-'), stringsAsFactors=F))
colnames(regions) = c('chr','start','stop')
regions[colnames(regions)] = lapply(regions[colnames(regions)], function(x) as.numeric(x))

@@

```{r, echo=T}
# positive coefficients were related to gains in progressors
gains = subset(ch, coef > 0)

# counts of gains per coef
gain = do.call(rbind, lapply(gains$label, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values > 1)])
}))
rownames(gain) = gains$label
wilcox.test(gain[,1], gain[,2])

@@

loss = do.call(rbind, lapply(losses$label, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values < -1)])
}))
rownames(loss) = losses$label
wilcox.test(loss[,1], loss[,2])
```

@@

randomGains = do.call(rbind, lapply(sample.reg, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values > 1)])
}))
wilcox.test(randomGains[,1], randomGains[,2])

randomLosses = do.call(rbind, lapply(sample.reg, function(reg) {
  values = rawSWGSdf[,reg] 
  table(labels[which(values < -1)])
}))
wilcox.test(randomLosses[,1], randomLosses[,2])
```

```{r echo=F}
rg = wilcox.test(gain[,'0'], randomGains[,'0'])
rl = wilcox.test(loss[,'0'], randomLosses[,'0'])
```
There appears to be no difference in gains (P=`r rg$p.value`) or losses (P=`r rl$p.value`) for Non-progressors as compared to random.

```{r echo=F}
rg = wilcox.test(gain[,'1'], randomGains[,'1'])
rl = wilcox.test(loss[,'1'], randomLosses[,'1'])
```
However, there is a significant difference in the selected regions for progressors as compared to random (gains P=`r rg$p.value` and losses P=`r rl$p.value`). This is mostly a sanity check.

@@

```{r, fig.height=16, fig.width=16}
plot.gains<-function(df) {
  df = lapply(df, function(x) {
    x = plyr::arrange(x, -width)
    x$order = factor(1:nrow(x))
    return(x)
  })

  m = do.call(rbind.data.frame, df)
  m = m %>% group_by(seqnames) %>% dplyr::ungroup() %>% 
    dplyr::arrange(segment, seqnames, width, start) %>% dplyr::mutate(segorder = row_number()) 
  m$isWGD = as.factor(m$isWGD)

@@

# WGD
tumorcnv$isWGD = tumorcnv$ploidy > 3.2 
becnv$isWGD = becnv$ploidy > 3.2

@@

## gains in patients -- issues with WGD samples?
gains = lapply(as(swgRegions[gr], "GRangesList"), gain.for, cnv=tumorcnv)
names(gains) = as.character(swgRegions[gr])
plot.gains(gains) + labs(title='Gains in OAC samples')

sequenced = base::merge(sequenced, demo, by.x='OCCAMS_ID', by.y='StudySubjectID')
subset(sequenced, RP.BarrettsAdjacent == 'yes')$Illumina_Barcode

be.cnvs = subset(tumorcnv, Tumor.Barcode %in%  subset(sequenced, RP.BarrettsAdjacent == 'yes')$Illumina_Barcode)
non.be.cnvs = subset(tumorcnv, Tumor.Barcode %in%  subset(sequenced, RP.BarrettsAdjacent == 'no')$Illumina_Barcode)

be.gains = lapply(as(swgRegions[gr], 'GRangesList'), gain.for, cnv=be.cnvs)
names(be.gains) = as.character(swgRegions[gr])
plot.gains(be.gains[sapply(be.gains, nrow) > 0]) + labs(title='Gains in BE Adjacent OAC samples')

non.be.gains = lapply(as(swgRegions[gr], 'GRangesList'), gain.for, cnv=non.be.cnvs)
names(non.be.gains) = as.character(swgRegions[gr])
plot.gains(non.be.gains[sapply(non.be.gains, nrow) > 0]) + labs(title='Gains in Non-BE Adjacent OAC samples')

@@

# Per patient
gfreq = sort(sapply(gains, function(x) length(unique(x$Illumina.Barcode)))/length(unique(sequenced$OCCAMS_ID)))
ggplot(melt(as.matrix(gfreq)), aes(Var1, value)) + geom_col(fill='darkblue') + ylim(0,0.5) + coord_flip() + 
  labs(x='', y='Frequency', title='Gain frequency in OAC') + theme_bw() + theme(text=element_text(size=14))

gfreq1 = sort(sapply(be.gains, function(x) length(unique(x$Illumina.Barcode)))/length(unique(be.cnvs$Illumina.Barcode)))
gfreq1 = gfreq1[names(gfreq)]

gfreq2 = sort(sapply(non.be.gains, function(x) length(unique(x$Illumina.Barcode)))/length(unique(non.be.cnvs$Illumina.Barcode)))
gfreq2 = gfreq2[names(gfreq)]

@@

```{r, fig.height=12, fig.width=12}
plot.losses<-function(df) {
  if (!is.data.frame(df)) {
    df = lapply(df, function(x) {
      if(nrow(x) <= 0) return(NULL)
      x = plyr::arrange(x, -width)
      x$order = factor(1:nrow(x))
      return(x)
    })
    m = do.call(rbind.data.frame, df)
  } else {
    m = df
    m = plyr::arrange(m, -width)
    m$order = factor(1:nrow(m))
  }
  #m$segnames = ordered(m$seqnames, levels=c(1:22))

  m = m %>% dplyr::group_by(seqnames) %>% ungroup() %>% dplyr::arrange(segment, seqnames, width, start) %>% dplyr::mutate(segorder = row_number())
  m$isWGD = as.factor(m$isWGD)
  m$CN = as.factor(m$Total_CN)
  #m = transform(m, segment=reorder(segment, segnames))
  p = ggplot(data=m, aes(x=order, y=start))

@@

losses = lapply(as(swgRegions[lr], 'GRangesList'), loss.for, tumorcnv)
names(losses) = as.character(swgRegions[lr])
plot.losses(losses)

be.losses = lapply(as(swgRegions[lr], 'GRangesList'), loss.for, be.cnvs)
names(be.losses) = as.character(swgRegions[lr])
plot.losses(be.losses) + labs(title="Losses in (YES) BE Adjacent samples")

non.be.losses = lapply(as(swgRegions[lr], 'GRangesList'), loss.for, non.be.cnvs)
names(non.be.losses) = as.character(swgRegions[lr])
plot.losses(non.be.losses) + labs(title='Losses in (NO) BE Adjacent samples')

@@

lfreq = sort(sapply(losses, function(x) length(unique(x$Illumina.Barcode)))/length(unique(sequenced$OCCAMS_ID)))
ggplot(melt(as.matrix(lfreq)), aes(Var1, value)) + geom_col(fill='darkgreen') + ylim(0,1) + coord_flip() + 
  labs(x='', y='Frequency', title='Loss frequency in OAC') + theme_bw() + theme(text=element_text(size=14))

lfreq1 = sapply(be.losses, function(x) length(unique(x$Illumina.Barcode)))/length(unique(be.cnvs$Illumina.Barcode))
lfreq1 = lfreq1[names(lfreq)]

lfreq2 = sort(sapply(non.be.losses, function(x) length(unique(x$Illumina.Barcode)))/length(unique(non.be.cnvs$Illumina.Barcode)))
lfreq2 = lfreq2[names(lfreq)]

@@

TODO: This should really separate the focal losses in 17p from the large-scale losses.

```{r fig.height=6,fig.width=8}
chr17 = get.chr.lengths()[17,]

chr17losses = losses[grep('^17:', names(losses))][[1]]

plot.losses(chr17losses) + geom_hline(data=chr17, aes(yintercept=(chr.cent-cent.gap)), color='red', linetype='dashed') +
  labs(y='chr17 position', x='Patient', title='17p Losses in (all) OAC Patients', subtitle=paste(round(length(unique(chr17losses$Illumina.Barcode))/length(unique(tumorcnv$Illumina.Barcode)), 2)*100, '% of OAC patients', sep=''))

@@

There does not appear to be a difference in gains vs losses.
```{r}
wilcox.test(sapply(gains, nrow), sapply(losses, nrow))
one.way.aov(sapply(gains, nrow), sapply(losses, nrow))
```

@@

These may be more nuanced, as we need to be aware of gains that reflect WGD vs gains that are new amplifications. At the moment there's no distinction.

There's no clear difference between the randomly selected regions and gains overall.  But there does appear to be a difference when comparing the randomly selected regions to the regions split by BE/non-BE.  I'm not sure this is real considering the overall case though.

```{r, echo=F}
randomGains = lapply(as(sample.regions,'GRangesList'), gain.for, cnv=tumorcnv)
names(randomGains) = as.character(sample.regions)
randomGains = randomGains[which(sapply(randomGains, nrow) > 2)]

@@

# Maybe a difference in # gains in the selected features vs random
# wilcox.test(sapply(randomGains, nrow), sapply(gains, nrow))
# 
# tests = matrix(ncol=2, nrow=10, dimnames=list(c(1:10), c('Rand.vs.BE.Gain', 'Rand.v.Non.Be.Gain')))
# for (i in 1:10) {
#   
#   sr = randomGains[sample(1:length(randomGains), length(be.gains))]
#   wt1 = wilcox.test(sapply(sr, nrow), sapply(be.gains,nrow))
#   
#   sr = randomGains[sample(1:length(randomGains), length(non.be.gains))]
#   wt2 = wilcox.test(sapply(sr, nrow), sapply(non.be.gains,nrow))
#   
#   tests[i,] = c(wt1$p.value, wt2$p.value)
# }
# tests < 0.005
# 
# #wilcox.test(sapply(randomGains, nrow), sapply(be.gains,nrow))
# #wilcox.test(sapply(randomGains, nrow), sapply(non.be.gains,nrow))
# 
# # Not a difference here though
# wilcox.test(sapply(gains, function(df) length(unique(df$Illumina.Barcode))), 
#             sapply(randomGains, function(df) length(unique(df$Illumina.Barcode))))

df = rbind(cbind.data.frame('counts'=sapply(gains, function(df) length(unique(df$Illumina.Barcode))), 'type'='selected') , 
      cbind.data.frame('counts'=sapply(randomGains, function(df) length(unique(df$Illumina.Barcode))), 'type'='random'))
ggplot(df, aes(type, counts, group=type, fill=type)) + geom_boxplot() + geom_jitter(width=0.3) + labs(title='Patient counts by region') + plot.theme

@@

```{r, echo=F}
randomLosses = lapply(as(sample.regions,'GRangesList'), loss.for, cnv=tumorcnv)
names(randomLosses) = as.character(sample.regions)

randomLosses = randomLosses[which(sapply(randomLosses, nrow) > 2)]

@@

# # No difference in # losses in the selected features
# wilcox.test(sapply(randomLosses, nrow), sapply(losses, nrow))
# 
# 
# tests = matrix(ncol=2, nrow=10, dimnames=list(c(1:10), c('Rand.vs.BE.Loss', 'Rand.v.Non.Be.Loss')))
# for (i in 1:10) {
#   sr = randomLosses[sample(1:length(randomLosses), length(be.losses))]
#   wt1 = wilcox.test(sapply(sr, nrow), sapply(be.losses,nrow))
#   
#   sr = randomLosses[sample(1:length(randomLosses), length(non.be.losses))]
#   wt2 = wilcox.test(sapply(sr, nrow), sapply(non.be.losses,nrow))
#   
#   tests[i,] = c(wt1$p.value, wt2$p.value)
# }
# tests < 0.005

# No difference between the two
df = rbind(cbind.data.frame('counts'=sapply(losses, function(df) length(unique(df$Illumina.Barcode))), 'type'='selected') , 
      cbind.data.frame('counts'=sapply(randomLosses, function(df) length(unique(df$Illumina.Barcode))), 'type'='random'))
ggplot(df, aes(type, counts, group=type, fill=type)) + geom_boxplot() + geom_jitter(width=0.3) + labs(title='Loss counts by patient') + plot.theme

@@

```{r, echo=F, fig.width=8}
size = unlist(sapply(gains, function(df) df$width))
hist(size)
abline(v = 3e6, col='red')

# Focal changes < 1% or the chromosome?
countBySize <-function(df) {
  s = round(subset(chr.lengths, chrom == unique(df$seqnames))$chr.length *.01)
  cbind('focal'=nrow(subset(df, width < s)),
        'large'=nrow(subset(df, width >= s)))
}

## Still in the regions selected
sizeGainsByPatient = do.call(rbind.data.frame,lapply(gains, sizebyPatient))
sizeGains = do.call(rbind.data.frame,lapply(gains, countBySize))

#  enrichment for some focal changes
x = which(p.adjust(apply(sizeGains, 1, function(x) binom.test(x[1], sum(x), 0.5)$p.value )) < 0.05)
# CI
apply(sizeGains, 1, function(x) binom.test(x[1], sum(x), 0.5)$conf.int )[,x]
